SVA Investigation

### 2021-09-27 Reports/FFModeling.Rmd
Reports/*.Rmdreport on ROC code

```{r bivariate table stratified by visit and ancestry child visit}
export2md(strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+pediatric+levine+anynewborn_center+SSnewbornCT_center+SSolder_center+Epi+Fib+IC+ChildAgeComposite+cm1inpov+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata%>%filter(childteen=='C'))), 'ancestry'))
```
## Methylation summary metrics, confounders and precision variables by prenatal maternal smoking, ancestry stratified at 15 year visit

>>>>

```{r bivariate table stratified by visit and ancestry child visit}
export2md(strataTable(createTable(compareGroups(smkPreg_binary~globalmethylation+pediatric+levine+anynewborn_center+SSnewbornCTw_center+SSolder_center+Epi+Fib+IC+ChildAgeComposite+cm1inpov+cm1bsex+m1g2_YesNoPreg+m1g3_YesNoPreg+PostnatalMaternalSmokingAny+SmkAtVisitPastmonth, data=modeldata%>%filter(childteen=='C'))), 'ancestry'))
```
## Methylation summary metrics, confounders and precision variables by prenatal maternal smoking, ancestry stratified at 15 year visit

### 2021-09-27 Reports/SVAReport.Rmd
Reports/*.Rmdreport on ROC code

---
title: "SVA Investigation"
author: "F blostein"
date: "9/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#libraries
library(sjlabelled)
library(tidyr)
library(dplyr)
library(purrr)
library(furrr)
library(stringr)
library(sva)
library(ggplot2)
```

@@

I tested the number of predicted surrogate varaibles at a variance filter of 1,000 (lowest recommended), 5,000, 10,000, 50,000 and 100,000 features. At 1,000 features there were 42 surrogate variables when using the be method. John said this is high in his experience. Increasing the number of features to 500 halves the number of surrogate variables detected to 20, after that there is a minor increase in SVs.  Using the leek method, I get 0 surrogate variables (except when using a variance filter of 50k, at which point I get 2 surrogate variables)

@@

```{r heatmap function}
bring.the.heatmap.sv <- function(pd,covar,folder){
  library(gplots)

@@

  #pdf(paste0(folder,'heatmap-sv.pdf'))
  color.gradient <- colorRampPalette(c('firebrick','ivory3','royalblue'),bias=3)
  heatmap.2(correlations,Rowv=FALSE,Colv=FALSE,cexRow=1,cexCol=1,dendrogram='none',density.info='none',trace='none',
            col=color.gradient(20), key.xlab='p-value', main='Associations Between SVs\n and Given Covariates',
            #lmat=rbind(c(0,3,3,3),c(2,0,4,0),c(0,1,1,1)), lhei=c(0.5,1.5,3), lwid=c(0.12,0.145,0.4,0.335),
            margin=c(4,12)
            )
  #dev.off()

@@

```{r plot heatmap, fig.width=20, fig.height=12}
bring.the.heatmap.sv(my.pd, c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Epi', 'Fib', 'IC', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
```

@@

```{r, echo=FALSE,results='hide',fig.keep='all'}
map(.x=pd.ancestry$pd, .f=~(bring.the.heatmap.sv(as.data.frame(.x), covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'cm1bsex', 'Epi', 'Fib', 'IC', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))))
```

@@

@@

global_models%>%unnest(tidied)%>%
  filter(term=='smkPreg_binaryYes')%>%
  #filter(outcome%in% my_outcomes)%>%
  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=predictors))+
  geom_point(position = position_dodge(width = 1))+
  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
  facet_grid(.~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)))+
  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
  theme_classic()+
  theme(text = element_text(size=16))+theme(legend.position = 'bottom')+
  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

```
## Remaining questions

>>>>

---
title: "SVA Investigation"
author: "F blostein"
date: "9/8/2021"
output:
    html_document:
      code_folding: show
      highlight: tango
      number_sections: yes
      theme: sandstone
      toc: yes
      toc_depth: 2
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)

#libraries
library(sjlabelled)
library(tidyr)
library(dplyr)
library(purrr)
library(furrr)
library(stringr)
library(sva)
library(ggplot2)
```

@@

I tested the number of predicted surrogate varaibles at a variance filter of 1,000 (lowest recommended), 5,000, 10,000, 50,000 and 100,000 features. At 1,000 features there were 42 surrogate variables when using the be method. John said this is high in his experience. Increasing the number of features to 5000 halves the number of surrogate variables detected to 20, after that there is a minor increase in SVs. In the full matrix SV analysis (using all the CpG sites), there are 36 syrrigate variables.  Using the leek method, I get 0 surrogate variables (except when using a variance filter of 50k, at which point I get 2 surrogate variables)

@@

```{r heatmap function}
bring.the.heatmap.sv <- function(pd,covar,folder, name='Associations Between SVs\n and Given Covariates'){
  library(gplots)

@@

  #pdf(paste0(folder,'heatmap-sv.pdf'))
  color.gradient <- colorRampPalette(c('firebrick','ivory3','royalblue'),bias=3)
  heatmap.2(correlations,Rowv=FALSE,Colv=FALSE,cexRow=1,cexCol=1,dendrogram='none',density.info='none',trace='none',
            col=color.gradient(20), key.xlab='p-value', main=name,
            #lmat=rbind(c(0,3,3,3),c(2,0,4,0),c(0,1,1,1)), lhei=c(0.5,1.5,3), lwid=c(0.12,0.145,0.4,0.335),
            margin=c(4,12)
            )
  #dev.off()

@@

```{r plot heatmap}
bring.the.heatmap.sv(my.pd, c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'ancestry', 'cm1bsex', 'Epi', 'Fib', 'IC', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'))
```

@@

```{r, echo=FALSE,results='hide',fig.keep='all'}
map2(.x=pd.ancestry$pd, .y=paste('SV and covariate \n associations in', pd.ancestry$ancestry),  .f=~(bring.the.heatmap.sv(as.data.frame(.x), covar=c('smkPreg_binary', 'Sample_Plate', 'Slide', 'Array', 'ChildAgeComposite', 'cm1bsex', 'Epi', 'Fib', 'IC', 'cm1inpov', 'm1g2_YesNoPreg', 'm1g3_YesNoPreg', 'PostnatalMaternalSmokingAny', 'SmkAtVisitPastmonth'), name=.y)))
```

@@

### Global w/ full matrix SVs

@@

global_models%>%unnest(tidied)%>%
  filter(term=='smkPreg_binaryYes')%>%
  #filter(outcome%in% my_outcomes)%>%
  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=predictors))+
  geom_point(position = position_dodge(width = 1))+
  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
  facet_wrap(.~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)), ncol=3)+
  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
  theme_classic()+
  theme(text = element_text(size=12))+theme(legend.position = 'bottom')+
  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')
```

### Local (ancestry-specific)

```{r}
local_models%>%unnest(tidied)%>%
  filter(term=='smkPreg_binaryYes')%>%
  filter(outcome%in% c('Global methylation', 'Pediatric clock', 'GRIM clock'))%>%
  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=predictors))+
  geom_point(position = position_dodge(width = 1))+
  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
  facet_grid(ancestry~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)))+
  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
  theme_classic()+
  theme(text = element_text(size=12))+theme(legend.position = 'bottom')+
  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')

local_models%>%unnest(tidied)%>%
  filter(term=='smkPreg_binaryYes')%>%
  filter(str_detect(outcome, 'score'))%>%
  ggplot(aes(x=estimate, xmin=conf.low, xmax=conf.high, color=factor(age, rev(levels(age))), y=predictors))+
  geom_point(position = position_dodge(width = 1))+
  geom_errorbarh(height=0.1, position = position_dodge(width = 1))+
  facet_grid(ancestry~outcome, scales='free_x', labeller = labeller(outcome=label_wrap_gen(15)))+
  geom_vline(xintercept=0, color='red', linetype="dashed", alpha=0.5)+
  theme_classic()+
  theme(text = element_text(size=12))+theme(legend.position = 'bottom')+
  scale_color_discrete('', limits=c('Age 9', 'Age 15'))+
  scale_x_continuous('Coefficient for prenatal maternal smoking + 95% CI')
```

## Remaining questions
